\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2016-05-01 doc/USER_MANUAL/01_introduction.tex
added a reference to the users manual

The SEM was originally developed in computational fluid dynamics \citep{Pat84,MaPa89}
and has been successfully adapted to address problems in seismic wave
propagation. Early seismic wave propagation applications of the SEM,
utilizing Legendre basis functions and a perfectly diagonal mass matrix,
include \citet{CoJoTo93}, \citet{Kom97}, \citet{FaMaPaQu97}, \citet{CaGa97},
\citet{KoVi98} and \citet{KoTr99}, whereas applications involving
Chebyshev basis functions and a non-diagonal mass matrix include \citet{SePr94},
\citet{PrCaSe94} and \citet{SePrPr95}.\\

@@

>>>>

The SEM was originally developed in computational fluid dynamics \citep{Pat84,MaPa89}
and has been successfully adapted to address problems in seismic wave
propagation. Early seismic wave propagation applications of the SEM,
utilizing Legendre basis functions and a perfectly diagonal mass matrix,
include \citet{CoJoTo93}, \citet{Kom97}, \citet{FaMaPaQu97}, \citet{CaGa97},
\citet{KoVi98} and \citet{KoTr99}, whereas applications involving
Chebyshev basis functions and a non-diagonal mass matrix include \citet{SePr94},
\citet{PrCaSe94} and \citet{SePrPr95}.
In the Legendre version that we use in SPECFEM the mass matrix is purposely slightly inexact but diagonal (but can be made exact if needed, see \cite{Teu15}),
while in the Chebyshev version it is exact but non diagonal.\\

@@

If you use the C-PML absorbing layer capabilities of the code, please cite at least one article
written by the developers of the package, for instance:
%
\begin{itemize}
\item \cite{XiKoMaMa14},
\item \cite{XiMaCrKoMa16}.
\end{itemize}
%
If you use the attenuation (anelastic/viscoelastic) capabilities of the code, please cite at least one article
written by the developers of the package, for instance:
%
\begin{itemize}
\item \cite{BlKoChLoXi16}.
\end{itemize}
%
If you use the kernel capabilities of the code, please cite at least one article
written by the developers of the package, for instance:
%
\begin{itemize}
\item \cite{TrKoLi08},
\item \cite{PeKoLuMaLeCaLeMaLiBlNiBaTr11},
\item \cite{LiTr06},
\item \cite{MoLuTr09}.
\end{itemize}

### 2016-05-01 doc/USER_MANUAL/05_running_the_solver.tex
fixed a typo

%------------------------------------------------------------------------------------------------%

%------------------------------------------------------------------------------------------------%

\chapter{Note on the SPECFEM acoustic formulation}

{\Large Note from Quentin Brissaud, Dimitri Komatitsch and Rapha\"el Garcia, October 2015}

\section{About the potential formulation used in SPECFEM for pure acoustic waves (i.e. in the absence of gravity terms)}

In SPECFEM, when considering purely acoustic waves (neglecting gravity acceleration and viscosity)
in fluid regions one uses a potential formulation to represent the "density times displacement" perturbation
\cite{ChKoViCaVaFe07}:
\begin{equation}
\rho \mathbf{u} = \nabla \chi
\label{potential}
\end{equation}
where $\mathbf{u}$ is the displacement perturbation, $\rho$ the background density and $\chi$ is the scalar potential.\\
This formulation has two great advantages: it reduces the number of unknowns from three to one, and it also automatically removes artefacts
in finite-element methods (FEM) caused by the displacement formulation \cite{HaOuVe78}.
This is crucial since it is the main reason for which a displacement formulation cannot be used in fluid regions in a FEM.
\newline

However, this formulation indirectly assumes that the density is spatially constant, or locally slowly varying (e.g. inside a given spectral element),
because acoustic displacements must be irrotational. Since we know that $\nabla \times \mathbf{u}$ must be equal to zero
we can see that when density has strong local variations, equation (\ref{potential}) is not valid any more.
Indeed, taking the rotational of \eqref{potential}
\begin{eqnarray*}
\nabla \times (\rho \mathbf{u}) &=& \nabla\times\nabla \chi \, ,
\end{eqnarray*}
since the curl of a gradient is always zero i.e. for any scalar $\chi$, $ \nabla\times\nabla \chi = 0$, we get
\begin{eqnarray*}
\nabla\rho\times\mathbf{u} + \rho\nabla\times\mathbf{u} &=& \mathbf{0}\\
\nabla\times\mathbf{u}&=&-(1/\rho)\nabla\rho\times\mathbf{u}
\end{eqnarray*}
Then, one notices that in the case of strong density gradients the rotational of $\mathbf{u}$ is no longer zero.
Thus, the potential formulation \eqref{potential} is no more valid. It is valid only when $(1/\rho)\nabla\rho\times\mathbf{u}$ can be considered negligible,
i.e. when both the gradient of density and the value of displacement are small, or if the displacement is parallel or quasi-parallel to the gradient of density,
in which case their rotational is also close to zero.
\newline

Another approach exists for acoustic waves, based on a displacement potential instead of on a potential of $\rho \mathbf{u}$:
\begin{equation}
\mathbf{u} = \nabla \chi \, ,
\end{equation}
which correctly gives $\nabla\times\mathbf{u} = \mathbf{0}$.
But in this case the gradient of density is neglected in the Eulerian momentum equation in order to identify gradient terms.
Indeed, from the Eulerian momentum equation for inviscid acoustic waves only
\begin{eqnarray*}
\rho\partial_t^2\mathbf{u} &=& -\nabla p\\
\nabla (\partial_t^2\chi)&=& -(1/\rho)\nabla p \, .
\end{eqnarray*}
When $\nabla\rho \gg \mathbf{0}$ strictly speaking one cannot identify $\partial_t^2\chi$ with $p$.

The solution to this is to work with pressure as the scalar unknown (\cite{God11}, equation 23),
or even better with the time integral (the primitive) of pressure (see e.g. \cite{Eve81});
this was the approach that was originally implemented in SPECFEM until 10 years ago or so, but \cite{ChCaVi03} showed that switching to a displacement potential
or to a potential of $\rho \mathbf{u}$ enables one to couple the fluid and solid regions without iterating on the coupling condition in the time scheme,
thus resulting in much easier and cheaper implementation; since at that time SPECFEM users were not studying media with strong local variations of density (in solid Earth
geophysics such large local variations in fluid regions never happen), the SPECFEM developers permanently switched to such a displacement-based potential formulation.
However, for media with strong local variations of density in a fluid region we should reconsider the issue and consider trying the pressure-based formulation again.

Cases in which this can be problematic may include e.g. upper atmosphere studies: the density there is very small but displacements can become very large;
in 1D atmosphere models $\frac{\nabla \rho}{\rho}$ is approximately constant and equal to -1/H, where H is the scaling height,
which leads to $\frac{\nabla \rho}{\rho} \simeq$ 3.42*1e-5, i.e. very small; however displacements there can be large
and thus it is not clear if $(1/\rho)\nabla\rho\times\mathbf{u}$ can be considered negligible there.

>>>>

\section{Note on the viscoelastic model used}

\noindent
The model used is a constant $Q$, thus with no dependence on frequency ($Q(f)$ = constant).
See e.g. \cite{BlKoChLoXi16}. \\

\noindent
However in practice for technical reasons it is approximated based on the sum of different Generalized Zener body mechanisms
and thus the code outputs the band in which the approximation is very good, outside of that range it can be less accurate.
The logarithmic center of that frequency band is the \texttt{ATTENUATION\_f0} parameter defined (in Hz) in input file \texttt{DATA/Par\_file}.

### 2016-05-01 doc/USER_MANUAL/08_doing_tomography.tex
made a sentence more precise in the users manual

Whatever misfit function you use for the tomographic inversion (you can find several
examples in \citet{TrKoLi08} and \citet{TrTaLi05}), you will weight
the sensitivity kernels with measurements. Furthermore, you will use
as many measurements (stations, components, time windows) as possible
per event; hence, we call these composite kernels ``event kernels'',
which are volumetric fields representing the gradient of the misfit
function with respect to one of the variables (\eg $V_{{\rm s}}$).
The basic features of an adjoint-based tomographic inversion were
illustrated in \citet{TrKoLi08} and \citet{TaLiTr07} using a conjugate-gradient algorithm.
Other, more powerful techniques such as the Limited-Memory Broyden-Fletcher-Goldfarb-Shanno (L-BFGS) method can now be used,
as illustrated for instance in \cite{MoChKoWa15}.

@@

There are dozens of versions of gradient-based inversion
algorithms that could alternatively be used. The tomographic inversion
of \citet{TaLiMaTr09,TaLiMaTr2010} used SPECFEM3D\_Cartesian as well
as several additional components which are also stored on the CIG
svn server, described next.

>>>>

Whatever misfit function you use for the tomographic inversion (you can find several
examples in \citet{TrKoLi08} and \citet{TrTaLi05}), you will weigh
the sensitivity kernels with measurements. Furthermore, you will use
as many measurements (stations, components, time windows) as possible
per event; hence, we call these composite kernels ``event kernels'',
which are volumetric fields representing the gradient of the misfit
function with respect to one of the variables (\eg $V_{{\rm s}}$).
The basic features of an adjoint-based tomographic inversion were
illustrated in \citet{TrKoLi08} and \citet{TaLiTr07} using a conjugate-gradient algorithm.
Other, more powerful techniques such as the Limited-Memory Broyden-Fletcher-Goldfarb-Shanno (L-BFGS) method can now be used,
as illustrated for instance in \cite{MoChKoWa15}.

@@

There are many other versions of gradient-based inversion
algorithms that could alternatively be used (see e.g. \cite{ViOp09,MoChKoWa15} for a list).
The tomographic inversion of \citet{TaLiMaTr09,TaLiMaTr2010} used SPECFEM3D\_Cartesian as well
as several additional components which are also stored on the CIG
svn server, described next.

### 2016-05-01 doc/USER_MANUAL/15_informations_for_developers.tex

\noindent We also have simple 3D demo source codes that implement the SEM in a single, small program, in directory\\
utils/small\_SEM\_solvers\_in\_Fortran\_and\_C\_without\_MPI\_to\_learn of the specfem3d package.
They are useful to learn how the spectral-element method works, and how to write or modify a code to implement it.
Also useful to test new ideas by modifying these simple codes to run some tests.\\

>>>>

\noindent We also have simple 3D demo source codes that implement the SEM in a single, small program, in directory\\
utils/small\_SEM\_solvers\_in\_Fortran\_and\_C\_without\_MPI\_to\_learn of the specfem3d package.
They are useful to learn how the spectral-element method works, and how to write or modify a code to implement it.
Also useful to test new ideas by modifying these simple codes to run some tests.
We also have a similar, even simpler, demo source code for the 2D case in directory\\
utils/small\_SEM\_solver\_in\_Fortran\_without\_MPI\_to\_learn of the specfem2d package.\\

### 2016-05-01 doc/USER_MANUAL/notes_and_acknowledgement.tex

\addcontentsline{toc}{chapter}{Notes \& Acknowledgments}

@@

\chapter*{Notes \& Acknowledgments}

@@

OpenDX \urlwithparentheses{http://www.opendx.org} is open-source
based on IBM Data Explorer, AVS \urlwithparentheses{http://www.avs.com}
is a trademark of Advanced Visualization Systems, and ParaView \urlwithparentheses{http://www.paraview.org}
is an open-source visualization platform.

>>>>

@@

\chapter*{Notes \& Acknowledgments}
\addcontentsline{toc}{chapter}{Notes \& Acknowledgments}

@@

OpenDX \urlwithparentheses{http://www.opendx.org} is open-source
based on IBM Data Explorer, AVS \urlwithparentheses{http://www.avs.com}
is a trademark of Advanced Visualization Systems, and ParaView \urlwithparentheses{http://www.paraview.org}
is an open-source visualization platform.\\
